---
title: 'Part 1: Getting Started'
author: "Matt Steele"
date: "`r Sys.Date()`"
output:
  html_document: 
   theme: cosmo
---

------------------------------------------------------------------------

## Resources

-   [Official Site for R and Comprehensive R Archive Network
    (CRAN)](https://cran.r-project.org/)

-   [Official Site for POSIT/RSTUDIO](https://rstudio.com/)

-   [O'Reilly Learning
    Platform](https://databases.lib.wvu.edu/connect/1540334373)

    -   R for Data Science, 2nd Edition

    -   R Programming for Statistics and Data Science
    
-   [Postit/RStudio Youtube Page](https://www.youtube.com/@PositPBC)

-   [R for Data Science Youtub Page](https://www.youtube.com/@r4ds)

------------------------------------------------------------------------

<br />

# Part 1: Getting Started


------------------------------------------------------------------------

## Why use R

-   It is open-source

    -   Free

    -   Platform independent

    -   Reproducible

    -   Shareable

    -   Contains add-on packages

-   Created for data statistical computation and graphic export

------------------------------------------------------------------------

## Create a File

You can create new R-focused files that allow you to document your code
and the outputs of your code.

-   **R Script**: create code based text file that allows you to save
    and execute code at your discretion

*File \> New File \> R Script*

-   **R Markdown** or **Quarto Document**: create literate code based
    files that allow you to write plain text language combined with your
    R code. This file outputs results from data analysis (including
    plots and tables) along with written commentary into a nicely
    formatted and reproducible document
*File \> New File \> R Markdown*

*File \> New File \> Quarto Document*

------------------------------------------------------------------------

## Set Your Preferences

*Tools \> Global Options*

Some suggested Preferences to set:

-   Code \> Editing \> Use Native Pipe Operator

-   Code \> Editing \> Soft wrap source R files

------------------------------------------------------------------------

<br />

## Overview RSTUDIO

RStudio is an Integrated Development Environment (IDE) that allows you
to save you code, store your variables and environments and view
outputs.

<br />

------------------------------------------------------------------------

**Console Pane**

pane on the left-hand side of the screen

-   This area is where your code is run and where outputs are displayed

------------------------------------------------------------------------

**Source Pane**

this pane is opened when you create or open a markdown or script file.

-   This area is where you can create code in script or markdown files
-   These files can be a saved and accessed at any point

------------------------------------------------------------------------

**Environment Pane**

view functions, objects, and data sets that are stored here

-   Your environment can be saved and accessed at any point
-   Save your environment to your working directory

------------------------------------------------------------------------

<br />

## Working Directory

The working directory in R is the folder where you are working. Hence,
it's the place (the environment) where you have to store your files of
your project in order to load them or where your R objects will be
saved.

<br />

------------------------------------------------------------------------

**Function: getwd()**

See the current directory you are in.

```{r}
#| label: getwd function
#| eval: false

getwd()

```

------------------------------------------------------------------------

**Function: setwd()**

You can set your working directory in RStudio by going to

*Session \> Set Working Directory \> Choose Directory*

-   Choose the folder you want to use for your current project

<br />

*HINT: After you set your working directory, save the path by copy and
pasting the file path from the console area into the source area using
the setwd() function*

<br />

```{r}
#| label: setwd function
#| eval: false

setwd("C:/Users/Matt/Documents/RWorkshop Development/workshop_r_for_marketing")

```

------------------------------------------------------------------------

<br />

# Part 2: Tips Before You Get Started


------------------------------------------------------------------------

## Keyboard Shortcuts

Since R is heavily focused on using code to execute commands it is
useful to become familiar with keyboard shortcuts

<br />

*Tools \> Keyboard Shortcuts Help*

<br />

### Popular Shortcuts:

|                         | PC                   | MAC             |
|-------------------------|:---------------------|:----------------|
| **Run Code**            | CTRL + ENTER         | CMD + RETURN    |
| **Assignment Operator** | ALT + -              | OPTION + -      |
| **Pipe Operator**       | CTRL + SHIFT + ENTER | CMD + SHIFT + M |

------------------------------------------------------------------------

## Commenting

Since use will be performing several operations in a single document and
even in a particular code chunk, it becomes important to document what
processes you were performing or make notes to useself or other about
your intentions.

You can do this in R using \*\*comments\*. Entering a hastag (\#) into
your code will comment anything that comes after for one single line.
Entering 3 single quotations (''') will comment out multiple lines.

```{r}
#| label: comments
#| eval: false

# get the mean of the mpg variable

mean(mtcars$mpg)

'''
The quick brown fox
jumps over
the lazy dog

'''

print("The quick brown fox jumps over the lazy dog")


```

------------------------------------------------------------------------

<br />

# Part 3: Coding Basics


------------------------------------------------------------------------

## Calculations

R can be used to perform simple and complex calculations

```{r}
#| label: calculations
#| eval: false

# Addition:
  129 + 3483
  
  # Subtraction:
  23693 - 4536
  
  # Multiplication:
  23 * 45
  
  # Division:
  51/3
  
  # Exponents:
  2^4
  
  # Logarithms (base e):
  log(100)


```

------------------------------------------------------------------------

## Built-in Functions and Arguments

<br>

**Functions:**

Like most computer software, R allows you to run commands. Commands in R
are referred to as functions. There are several built-in functions in
base R, however, when you install and call a new package, you will have
access to more functions that you can use.

```{r}
#| label: example function
#| eval: false

sample(1:5000)
```

<br>

**Arguments:**

Arguments are the values and parameters that are acted on by the
function.

```{r}
#| label: example arguments
#| eval: false

sample(1:5000, size = 50, replace = TRUE)

# arguments in a function have an order. Entering information in the order will allow you to skip the argument name

sample(1:5000, 50, TRUE)

```

------------------------------------------------------------------------

## Learn more about Functions and Arguments

<br>

-   [Base R Function
    Documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html)
-   For every package you want to use in R, there will be a
    documentation website or document. Explore these documents to see
    the available functions and their default arguments
-   Get help with a package, function, or argument within R. Using
    **(?)** or **help("**name of function")
    
<br>

```{r}
#| label: function and package help
#| eval: false

?dplyr

help("mean")

```

<br>

-   The **TAB** will auto-suggest options for you in R, when typing a
    name of a function. Scrolling over the name of the of the function
    will give you an overview of what is does and the default/available
    arguments of the function.

<br>

```{r}
#| label: tab button
#| eval: false

dplyr::
```

------------------------------------------------------------------------

## Objects

Objects allow you to store information for future recall

<br>

|                                |                                      |
|:-------------------------------|:-------------------------------------|
| numeric value                  | numValue \<- 400                     |
| character                      | chrValue \<- "Hello World"           |
| results of running function(s) | resultFunction \<- mean(mtcars\$mpg) |
| vector                         | vecValue \<- c(1,2,3)                |
| data frame                     | r dfValue \<- read.csv("mtcars.csv") |

: **Examples**

<br>

### Assignment Operator

The assignment operator (\<-) allows you to create an object.

|                         | PC      | MAC     |
|-------------------------|:--------|:--------|
| **Assignment Operator** | ALT + - | CMD + - |
: **Keyboard Shortcut**

<br>

```{r}
#| label: example object
#| eval: false

a <- 35
b <- 45

a
a + b
```

------------------------------------------------------------------------

### Naming Conventions

Objects should start with a *lowercase letter* and should not start with
a capital letter. Ideally, you should not name your object with the name
of a function, to avoid confusion.

*exampleOne \<- 45*

*example.two \<- mean(45)*

*example_three \<- sample(1:500, 50)*

------------------------------------------------------------------------

## Common Data Values

<br>

|               |                                                                                              |
|--------------------|----------------------------------------------------|
| **Double**    | regular numbers (large small, positive, negative, with digits after the decimal, or without) |
| **Integer**   | positive whole number with nothing after the decimal. Denoted using **( L )**                |
| **Character** | non-numeric data to be interpreted as text. Denoted using **( "" )**                         |
| **Factor**    | data objects which are used to categorize the data and store it as levels.                   |
| **Boolean**   | boolean data objects that denote TRUE or FALSE. Denoted with capital letters.                |

<br>

```{r}
#| label: data values
#| eval: false

# double

str(400)

# integer

str(400L)

# character

str("Hello World")

# factor

str(as.factor(mtcars$cyl))

# boolean

str(TRUE)
```

------------------------------------------------------------------------

## Packages

R is open source code after it's initial development people began adding
to it with packages. An R Package is something that you can plug into
RStudio to extend the basic functionality that is built in with R. One
of the reasons that R has become so popular is because it has this rich
ecosystem of packages that really make R a comprehensive platform for
data science.

---

### Install Packages

You must install a package before you can load it. *But you only need to
install it one time.*

```{r}
#| label: install packages
#| eval: false

install.packages("tidyverse")

```

<br>

### Load Packages

For every new session, you must load it to use the package's functions.

```{r}
#| label: load package
#| eval: false

library(tidyverse)

```

<br>

------------------------------------------------------------------------

<br>

# Part 4: Entering Data


------------------------------------------------------------------------

## Vectors

one-dimensional sequence of data elements. Data must be of the same
value.

<br>

*vectorSyntax \<- c(object1, object2, object3)*

<br>

```{r}
#| label: example vectors
#| eval: false

vec_one <- c(1,2,3)
vec_two <- c(4:6)

vec_two

```

<br>

---

## Lists

one-dimensional sequence of data elements. Data can contain multiple classes of
values.

<br>

*listSyntax \<- list(object1, object2, object3)*

<br>

```{r}
#| label: list example
#| eval: false

list_example <- list("hello", "goodbye", 4)

```

<br>

---

## Matrix

two-dimensional sequence of data elements. Data must be of the same
value.

<br>

*matrixSyntax \<- matrix(c(vector1, vector2))*

<br>

```{r}
#| label: matrix example
#| eval: false

matrix_example <- matrix(c(vec_one, vec_two), ncol = 2)
matrix_example
```
<br>

---

## Data Frame

two-dimensional sequence of data elements. Data can contain multiple classes of
values. Variables must be same values.

<br>

*dataFrameSyntax <- data.frame(column1, column2, column3)*


<br>

```{r}
#| label: create data frame
#| eval: false

# create vectors
title <- c("Star Wars", "The Empire Strikes Back", "Return of the Jedi")
year <- c(1972, 1980, 1983)
length.min <- c(121, 124, 133)
box.office.mil <- c(787, 534, 572)
  
# combine these vectors with the data.frame() function

  starWars.data <- data.frame(title, year, length.min, box.office.mil)
  starWars.data

```

<br>

---

<br>

# Part 5: Working with Data Frames


---

## Subsetting Variables

look at and/or work on individual variable in a data frame.

<br>

*object_name$variable_name*

<br>

```{r}
#| label: subsetting
#| eval: false

starWars.data$year
```

<br>

---

## Indexing Observations

Indexing is a way that you can call a column, row, or specific position in a data frame. You can use indexing to pinpoint a specific value and change that value.

<br>

*object_name[rownumber, columnnumber]*

<br>

```{r}
#| label: indexing
#| eval: false

starWars.data [1, 2] # view specific variable
starWars.data [1, ] # view observation across all variables
starWars.data [ , 2] # view observation in a variable

starWars.data [1, 2] = "1977" # change an observation

```
<br>

---

## Append New Variables

Create new variables with observations and append them to your data frame.

<br>

*cbind(vectorOne, vectorTwo)*

<br>

```{r}
#| label: append new varaibles with cbind 
#| eval: false

han <-  c(18.7, 21.4, 16.3) # create an new vector of time on screen for han solo

starWars.data <- cbind(starWars.data, han)
starWars.data

```

<br>

---

## Merge New Observations

Create new observations and merge it into a data frame

<br>

*rbind(objectOne, objectTwo)*

<br>

```{r}
#| label: merge new observations
#| eval: false

solo <- list("Solo", 2018, 135, 393.2, 55.5)

starWars.data <- rbind(starWars.data, solo)
starWars.data

```

<br>

---

## Export the Data Frame

Once you are done entering your data, you can export it to your working directory. The function without built-in arguments is **write.table( )** but if are saving it as a csv, you are better using **write.csv( )**.

<br >

*write.csv(object, "name of file.extension")*

<br>

```{r}
#| label: export files
#| eval: false

write.csv(starWars.data, "starwars.csv")
```

---

<br>

# Part 6: Explore Data 


---

<br>

```{r}
#| label: load tidyverse
#| eval: false

# if you have not yet loaded tidyverse let's go ahead and do that

library(tidyverse)


```

## Read a Data Frame into R

load data from a file in your working directory using the **read_csv( )** function from tidyverse. There is also the **read.csv( )** function in base R.

<br>

*read_csv("name of file.extension)*

<br>

```{r}
#| label: read data
#| eval: false

fight_songs <- read_csv("fight-songs.csv")

```

---

## Explore the Data Frame

<br>

---

### Function: view()

view the full data frame

<br>

```{r}
#| label: view
#| eval: false

view(fight_songs)
```

<br>

### Function: nrow()

view the number of rows/observations

<br>

```{r nrow, eval =FALSE}

    nrow(fight_songs)

```

<br>

### Function: ncol() 

view the number of columns/variables

<br>
    
```{r ncol, eval=FALSE}


  ncol(fight_songs)

```
  
### Function: colnames()

view the names of columns/variables

<br>
    
```{r colnames, eval=FALSE}

  colnames(fight_songs)

```

### Function: str()

view the structure of data

<br>
    
```{r structure, eval=FALSE}

  str(fight_songs)

```

### Function: summary()

basic descriptive statistics

<br>
    
```{r summary, eval=FALSE}


  summary(fight_songs)

```

### Function: mean(), median(), mode(), count(), sum() 

individual descriptive statistics

<br>
  
```{r descriptive statistics, eval=FALSE}

sum(fight_songs$number_fights)

```

### Function: head() or tail()

view top and bottom of data

<br>
    
```{r head and tail, eval=FALSE}


  head(fight_songs)

```

---
