---
title: "Part 3: Data Visualization"
author: "Matt Steele"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
---

<br />

------------------------------------------------------------------------

# Part 1: Introduction to GGPLOT2

visualize your data with ggplot

-   [ggplot package]("https://ggplot2.tidyverse.org/")


## Grammar of Graphics

ggplot allows you to visualize data by using three components: *a data set*, *a coordinate systems*, and *geoms*.

1.    Set the plot
2.    Set the coordinates
3.    Set the shape

<br />

------------------------------------------------------------------------

# Part 2: Setting the Plot and Coordinates

## Parameters

in ggplot you set you parameters or the variables you want to plot with
the **aesthetics** argument.

```{r}
#| label: aes
#| eval: false

# filter covid19 data frame to West Virginia

covid19_wv <- covid19_df |> 
  filter(State == "West Virginia") |> 
  group_by(Date) |> 
  summarise(Total = sum(Total))

covid19_wv


# set the parameters

covid_wv.line <- ggplot(covid19_wv, aes(Date, Total))
covid_wv.line


```

<br />

------------------------------------------------------------------------

# Part 3: Setting the plot shape

the **geom\_** or **stat\_** functions allows you to choose the shape.
We are going start by plotting a line graph so I will use
**geom_line()**.

```         
We combine elements in ggplot using the (+) operator 
```

```{r}
#| label: geom_line
#| eval: false

covid_wv.line +
  geom_line()
  

```

<br />

------------------------------------------------------------------------

# Part 4: Adding additional features

<br />

---

## Change color, size, and transparency of elements

|                           |                     |
|---------------------------|---------------------|
| **fill =** or **color =** | change colors       |
| **size =**                | change size         |
| **alpha =**               | change transparency |

```{r}
#| label: color, shape, transparency
#| eval: false

# add color

covid_wv.line +
  geom_line(color = "skyblue", show.legend = FALSE, size = 3, alpha = .75) +
  theme_classic()

# you can also adjust the background with built-in themes


```

<br />

---


## Labeling your plot

the labs element will allow you to add or change labels in the plot

```{r}
#| label: labs
#| eval: false

covid_wv.line + 
  geom_line(color = "skyblue", show.legend = FALSE, size = 3, alpha = .75) +
  theme_classic() + 
  labs( 
    title = "COVID19 Count", 
    subtitle = "West Virginia | 2020", 
    x = NULL,
    y = NULL )


```

<br />

---

## Change date elements

the **scale_x_date** element allows you to adjust your date elements on the x axis

[Date Formats - strptime](https://rdrr.io/r/base/strptime.html)

```{r}
#| label: scale_x_date
#| eval: false

covid_wv.line + 
  geom_line(color = "skyblue", show.legend = FALSE, size = 3, alpha = .75) +
  theme_classic() + 
  labs( 
    title = "COVID19 Count", 
    subtitle = "West Virginia | 2020", 
    x = NULL,
    y = NULL ) +
  scale_x_date(date_breaks = "3 weeks", 
               date_labels = "%b %d" )
  
```
<br />

---

## Creating a Histogram


```{r}
#| label: histogram
#| eval: false

# create an new data frame

customer_df <- customer_replaceNA |> 
  drop_na()

# create a histogram

customer.hist <- ggplot(customer_df, aes(Age))

customer.hist +
  geom_histogram()

# bin and add some color

customer.hist +
  geom_histogram(bins = 30, color = "coral", fill = "#ffbf00", alpha = .5)


```

<br />

---

## Adjust additional elements in the plot with theme

the **theme** element allows you adjust additional aspects of your plot. In this example we will change the font size and colors of the axis text.


```{r}
#| label: theme options
#| eval: false

# using the theme options

customer.hist +
  geom_histogram(bins = 30, color = "coral", fill = "#ffbf00", alpha = .5) +
  theme_light() +
  theme(axis.text = element_text(size = 4), axis.title = element_text(color = "#097969", size = 14)) +
  labs(
    y = NULL
  ) 

```

<br />

---

## Add a reference line with geom_vline or geom_hline

```{r}
#| label: vline
#| eval: false

# add a reference line

customer.hist +
  geom_vline(xintercept = 55, color = "#afe1af", size = 5) +
  geom_histogram(bins = 30, color = "coral", fill = "#ffbf00", alpha = .9) +
  theme_light() +
  theme(axis.text = element_text(size = 4), axis.title = element_text(color = "#097969", size = 14)) +
  labs(
    y = NULL
  ) 

```


<br />

---

## Seperate by categorical variables with facets

```{r}
#| label: facets
#| eval: false


# separate by categorical data with facets

customer.hist +
  geom_histogram(bins = 30, color = "coral", fill = "#ffbf00", alpha = .9) +
  theme_light() +
  theme(axis.text = element_text(size = 4), axis.title = element_text(color = "#097969", size = 14)) +
  labs(
    y = NULL
  ) + 
  facet_wrap(facets = vars(Department_Name), ncol = 3)

```

<br />

------------------------------------------------------------------------

# Bar Graph

```{r}
#| label: bar graph
#| eval: false

customer_afinn <- customer_sentiment |> 
  group_by(word) |> 
  summarise(n = sum(n), afinn_value = sum(afinn_value))


arrange(customer_afinn, desc(afinn_value))


customer_afinn <- customer_afinn |> 
  filter(afinn_value <= -600 | afinn_value >= 6000) |> 
  inner_join(bing)

customer_afinn

# you can order the bar graph using the fct_reorder function from Forcats

customer.bar <- ggplot(customer_afinn, aes(fct_reorder(word, afinn_value), afinn_value))

# you can color plots by categorical variables using the fill or color option followed by the variables name in the aes() argument

customer.bar + 
  geom_col(aes(fill = sentiment)) +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired")

# difference between geom_bar and geom_col

customer_df

customer.bar2 <- ggplot(customer_df, aes(Department_Name))

customer.bar2 +
  geom_bar()

customer.bar3 <- ggplot(customer_df, aes(Department_Name, Rating))

customer.bar3 +
  geom_col()

```
